Understanding Structure and Selectivity in Nanostructured Lyotropic Liquid Crystal Membranes using Molecular Simulations

### 2020-04-10 Ben_Manuscripts/thesis/thesis.tex
Shortened abstract

  The ability to perform highly selective aqueous separations at low cost
  is beneficial to a variety of chemical industries. Periodic, nanostructured 
  porous polymer membranes made from the cross-linked inverted hexagonal phase
  of self-assembled lyotropic liquid crystals (LLCs) are a promising class of 
  materials for selective separations. In this thesis, we use atomistic molecular 
  modeling in order investigate an experimentally characterized LLC polymer 
  membrane.

  First, we gain a deep understanding of nanoscopic pore structure in order 
  to make sure we can appropriately model the chemical environment in the pores.
  In particular, we compare simulated X-ray diffraction (XRD) patterns with 
  experimental XRD data to quantify and understand the major structural
  features that we must incorporate into our model. We explore the composition 
  and structure of the nanopores and reveal that there exists a composition gradient
  rather than an abrupt partition between the hydrophilic and hydrophobic regions.
  A caveat is that the time scales of the dynamics are extremely long for this system,
  resulting in simulated structures that appear too ordered, thus requiring
  careful examination of the metastable states observed in order to draw any
  conclusions.

  We use our carefully constructed molecular model in order to observe transport of
  water, sodium ions and 20 small polar solutes within the pores of a lyotropic 
  liquid crystal (LLC) membrane using atomistic molecular simulations. We find 
  that the transport of a species is dependent not only on molecular size, but on
  chemical functionality as well. We observe that all solutes perform intermittent 
  hops between lengthy periods of entrapment. Three different trapping mechanisms 
  are responsible for this behavior. First, solutes that drift out of the pore can
  become entangled among the dense monomer tails. Second, solutes can donate 
  hydrogen bonds to the monomer head groups. Third, solutes can coordinate with 
  sodium counter ions. The degree to which a solute is affected by each mechanism 
  is dependent on the chemical functionality of the solute.

  The remainder of this thesis focuses on developing stochastic models meant to 
  generate realizations of solute trajectories with similar qualitative and
  quantitative dynamics to our MD trajectories solute trajectories. Mathematically 
  modeling complex transport phenomena at the molecular level can be a powerful tool
  for connecting microscopic chemically-dependent solute motions that do not
  follow simple diffusive behavior with macroscopic membrane performances. 
  We develop three different stochastic time series models, parameterized by long 
  molecular dynamics (MD) simulation trajectories.

  First, using anomalous diffusion theory, we show how solute dynamics can be
  modeled as a fractional diffusion process subordinate to a continuous time
  random walk. From our MD simulations, we parameterize the distribution of
  dwell times, hop lengths between dwells and correlation between hops.

  Next, we generalize Markov state models, treating the configurational states of 
  the system as a Markov process where each state has distinct transport properties. 
  For each state and transition between states, we parameterize the distribution 
  and temporal correlation structure of positional fluctuations as a means of 
  characterization and to allow us to predict solute MSDs.

  Building on what we learned from the first two models, we develop a third approach,
  called the infinite hidden Markov model, which is capable of automatically 
  distinguishing an unknown number of hidden dynamical modes in multidimensional
  solute time series. This model offers a flexibile approach to modeling solute
  motion and helps us to uncover mechanisms with minimal human effort.
  }

>>>>

  Periodic, nanostructured porous polymer membranes made from the cross-linked
  inverted hexagonal phase of self-assembled lyotropic liquid crystals (LLCs) are
  a promising class of materials for selective separations. Our current understanding
  of these systems is not sufficient to design them for specific applications. In 
  this thesis, we use detailed atomistic molecular modeling in order to investigate 
  structure and transport of solutes within LLC polymer membrane.

  We begin by enhancing our understanding of the nanoscopic pore structure in order 
  to ensure we appropriately model the chemical environment in the pores. We use radial
  distribution functions and simulated X-ray diffraction patterns to quantify and 
  understand the major structural features of the model.

  We use our carefully constructed molecular model in order to observe transport of
  water, sodium ions and 20 small polar solutes within the pores of a lyotropic 
  liquid crystal (LLC) membrane using atomistic molecular simulations. We 
  observe that solutes perform intermittent hops between lengthy periods of 
  entrapment and describe how transport of a species is dependent not only on 
  molecular size, but on chemical functionality as well.

  Based on our qualitative understanding of solute transport, we focus the remainder
  of our efforts on the development of three different stochastic models which can be
  used to generate realizations of solute trajectories with qualitative and quantitative 
  dynamics similar to our MD solute trajectories. Mathematically modeling complex 
  transport phenomena at the molecular level can be a powerful tool for connecting 
  microscopic chemically-dependent solute motions that do not follow simple diffusive
  behavior with macroscopic membrane performance.

  First, using anomalous diffusion theory, we show how solute dynamics can be
  modeled as a fractional diffusion process subordinate to a continuous time
  random walk. We parameterize the distribution of dwell times, hop lengths between
  dwells and correlation between hops. Next, we generalize Markov state models, treating
  the configurational states of the system as a Markov process where each state has 
  distinct transport properties. Finally we implement a third approach,
  called the infinite hidden Markov model, which is capable of automatically 
  distinguishing an unknown number of hidden dynamical modes in multidimensional
  solute time series. This model offers a flexibile approach to modeling solute
  motion and helps us to uncover mechanisms with minimal human effort.
  }
